---
title: "Titanic Dataset Exploration"
author: "Sangeeta"
date: "16 july 2017"
output:
  html_notebook: default
  html_document: default
  pdf_document: default
---

#Question :  
What kind of people survived for example kids,women,men 


#Acquire Data 

As the problem statement itself carring the two set of Data training data,test data
Here we will just have a look of what kind of data it is .


```{r} 

library(dplyr)
library(knitr)
library(DT)


train_df<-read.csv("Titanic_DataSet/train.csv",stringsAsFactors=F)
cat(paste('\ntest data structure having rows :\n', nrow(train_df)) )
str(train_df)


test_df<-read.csv("Titanic_DataSet/test.csv",stringsAsFactors=F)
cat(paste('\ntest data structure having rows :\n', nrow(test_df)) )
str(test_df)




datatable(train_df,options=list(pageLength=5))


cat('\n test_df ----\n')


datatable(test_df,options=list(pageLength=5))



```
Insights:

1. As from the above result we can see that training Data structure is having  one field extra that is "survived" as comapred to       testing data which we have to predict in this data 

2. PasseengerId:Unique Id Assigned to each traveller

3. Pclass : 1 for First Class
            2 for Second Class
            3 for third class
            
4.  Name : Name is with first name and Last name Mostly starting from the Mr.Miss And Mrs. Few caseswith Master and don etc .

5.  Age :  Age of Passenger, few values are missing here

6.  SibSp : Number of sibling or spouse 0 for none 

7.  Parch : Number of parents/childern travelling  0 for none 

8.  Ticket : Ticket Number 

9.  Fare : how much each traveller paid for the travel

10. Cabin : Cabin number allocated

11. Embarked: Port of emabarkation C,S,Q 
            
             
#Explore

Now we will start exploring the data 

First we will do Univariate analysis,which means finding the continous and categorical variable
Continous Variable are : Passenger Id, PClass, Age,Sibsp,Parch,Ticekt Number ,Fare
Categorical : Sex, Embarked 
Character value Variable : Name ,Cabin,Ticekt 


Get here the summary of relavant data like,pclass,age,sibsp,parch,fare to see the average  

```{r}


library(ggplot2)

library(plotly)
library(grid)
library(gridExtra)

cat('\nSummary Of Pclass\n')

par(mfrow=c(1,2))

#Summary for Pclass
ggplotly (ggplot(train_df, aes(x="",y = Pclass)) +
    geom_boxplot(colour = "red", fill = "skyblue")+ggtitle("Pclass Summary"))

#Summary for Age 
ggplotly (ggplot(train_df, aes(x="",y = Age)) +
    geom_boxplot(colour = "red", fill = "skyblue")+ggtitle("Age Summary"))

#Summary for SibSp

ggplotly (ggplot(train_df, aes(x="",y = SibSp)) +
    geom_boxplot(colour = "red", fill = "skyblue")+ggtitle("SibSp Summary"))


#Summary for Parch
ggplotly (ggplot(train_df, aes(x="",y = Parch)) +
    geom_boxplot(colour = "red", fill = "skyblue")+ggtitle("Parch Summary"))

#Summary for Fare

ggplotly (ggplot(train_df, aes(x="",y = Fare)) +
    geom_boxplot(colour = "red", fill = "skyblue")+ggtitle("Fare Summary"))



```
Insight:

From the Results few things can be identified 

1. Average age of traveller is around 29-30 and maximum age is 80

2. very few people were traveling with siblings and maximum number of siblings/spouse is 8

3.  Maximum number of parents/childern is 6 and more outl

4. Average Fare is around 14.45 but outliers are more so prediction from fare will be tough


let see,from the given if we can find the Fare of first class, second class and third class
Not sure it is required or not.

```{r}


fare_single_passenger<-subset(train_df,(train_df$SibSp == 0) & (train_df$Parch ==0),select=c(Pclass,Fare))

fare_single_passenger

datatable(fare_single_passenger,options=list(pageLength=5))

fare_single_passenger<-aggregate(fare_single_passenger[, 2], list(fare_single_passenger$Pclass), mean)

           
cat(paste('\n\nAverage Fare of First Class ',fare_single_passenger[1,2]))
cat(paste('\n\nAverage Fare of Second Class ',fare_single_passenger[2,2]))
cat(paste('\n\nAverage Fare of Third Class ',fare_single_passenger[3,2]))





```

Change the variable into numeric  to fit on continous scale 

```{r}

train_df$Sex<-as.numeric(as.factor(train_df$Sex))
train_df$Embarked<-as.numeric(as.factor(train_df$Embarked))

test_df$Sex<-as.numeric(as.factor(test_df$Sex))
test_df$Embarked<-as.numeric(as.factor(test_df$Embarked))


```
further will visualize the data between survival and other variates .

here i am using the bar plot and manipulate so that user can choose the column value,number of row  and see the respective bar plot

```{r}

library(ggplot2)#library loadedr
library(ggthemes)

library(manipulate)




############# manipulate for the size and aes value


manipulate(
    ggplot(data=train_df[sample(1:nrow(train_df),samplesize),], aes_string(x = x_axis, y = "Survived",fill=x_axis,size=samplesize)) +
        geom_bar(stat="identity",width=0.5)+
        scale_fill_gradient(low="blue", high="red")+
        theme_solarized() ,
    samplesize=slider(100,nrow(train_df)),
    x_axis=picker("Pclass","Age","Sex","SibSp","Parch","Embarked")
    
)
###################



```
Insights:

1.From the above few points number female who survived is more as compared to male 

2. Can't predict much from the bar plot as there is no significant relationship seen



Check the survival of group/family  travelling together and singleton here i will use the ticket Number as if people are sharing the same ticket Number ,more chances that they are family or group travelling together .


```{r}

library(DT)

lenofUniqueTicekt<-length(unique(train_df$Ticket))

cat(paste('Number of uniques Ticket',lenofUniqueTicekt))


FamGp_df<-train_df[duplicated(train_df[,"Ticket"])|duplicated(train_df[,"Ticket"],fromLast  = TRUE),] %>%
          arrange(Ticket) %>%group_by(Ticket) %>% summarise(groupFmSize=n(),Survived=sum(Survived))




datatable(FamGp_df, options = list(pageLength = 5))



```
From the above table looks like possiblity of survival is little better  if travelling with group/Family 

before going into more visulaization just wanted to check which variable is playing a significant role in predicticing the survival applying logistics regression as the dependent variable is in binary form (0,1 ) means survived or not .
Removing few columns like name passengerID ticket,cabin as name and passenger ID is unique and cabin is having lots of missing data .Ticket is again character value not worth .

```{r}

training_model <- glm(Survived ~.-(Name+PassengerId +Ticket+Cabin),family=binomial(link='logit'),data=train_df,maxit=50)
summary(training_model)

```
From the above result we can see the following

1.SibSp, Fare and Embarked are not statistically significant.
2.Sex has lowest p-Value indicating strong association between the gender and survival rate 


Hence taking few Few more visualization related to Pclass,Sex,Age using plotly 

```{r}
library(ggplot2)
library(plotly)
#Draw thw chart for Male and female survivela

p<-ggplot(train_df, aes(Age, fill = factor(Survived))) + 
  geom_histogram() + 
  facet_grid(.~Sex)+theme_dark()

ggplotly(p)

```
```{r}

library(crosstalk)

shared_train_df <- SharedData$new(train_df)

bscols(d3scatter(shared_train_df,~SibSp,~Age,~Sex,width="100%",height = 300),d3scatter(shared_train_df,~Parch,~Age,~Sex,width = "100%",height = 300)
    )


```
Insights

Clearly from the graph that female of age between 12-42 survived and very few male survived between the same age group 
so prediction make better add few more column like child ,FemaleAdult ,maleAdult category as they have Female  adult high possiblity of survival.



Here on adding  both the data(training and test ), as need to filled the missing values in both  .

```{r}
#combine both the data 

library(dplyr)

titanic_df<-bind_rows(train_df,test_df)

#check the totral number of rows just to verify 
cat(paste('\ncomplete  data structure having rows :', nrow(titanic_df)) )
#checking the structure as well
str(titanic_df)




```

Add the Family size to each person 

```{r}
library(dplyr)
library(DT)

nrow((titanic_df))


#again recreate the Family/Group Size with whole data as we are going to add the family size in the test data to make prediction better 
FamGp_df<-titanic_df[duplicated(titanic_df[,"Ticket"])|duplicated(titanic_df[,"Ticket"],fromLast  = TRUE),] %>%
          arrange(Ticket) %>%group_by(Ticket) %>% summarise(groupFmSize=n(),Survived=sum(Survived))


#keep only two column Ticket ,groupFmSize  
FamGp_df <-FamGp_df[,c("Ticket","groupFmSize")]  

#display the data



datatable(FamGp_df, options = list(pageLength = 5))



```
From the above we have got family/group size 
which we will merge the groupFmSize to get the size in the whole


```{r}

library(dplyr)

titanic_df<-merge(x=titanic_df,y=FamGp_df,by="Ticket",all=TRUE)

#just to cross verify check the number of rows

cat(paste('\ncomplete  data structure having rows :', nrow(titanic_df)) )



#reorder the column value

col_order<-c("PassengerId","Survived","Pclass","Name","Sex","Age","SibSp","Parch","Ticket","groupFmSize","Fare","Cabin","Embarked")

titanic_df<-titanic_df[,col_order] %>% arrange(PassengerId)





datatable(titanic_df, options = list(pageLength = 5))



```
Replacing all the NA values in groupFamSize  with 0 

```{r}
titanic_df$groupFmSize[is.na(titanic_df$groupFmSize)]<-0

datatable(titanic_df, options = list(pageLength = 5))
```

Insights:

From the above we got complete data Now let us analyze which all value are missing 

No missing value in  data passenserId,Name,Sex,SibSp,Parch ,Ticket,Pclass
Few Missing value in Age , Cabin and two missing value in Embarked and one missing value in fare


here i am ignoring few of attributes which is not much useful as we saw earlier like embarked,cabin(having more missing value) and fare as well 


check the how many data is missing  using the mice plot
```{r}


 library(VIM)
library(plotly)


 mice_plot <- aggr(titanic_df[,c("Age")], col=c('blue','red'),
                    numbers=TRUE, sortVars=TRUE,
                    labels=c("Age")
                   , cex.axis=.7,
                    gap=3, ylab=c("Missing data","Pattern"))
 
```
Install Package install.packages("Hmisc")

filling  the missing value in Age with the impute method taking the meadin of the data

```{r}
library(Hmisc)
titanic_df$Age <- with(titanic_df, impute(Age, median))
#titanic_df[,c("Age")]

```
now all the missing value in age has been filled 

As we discussed earlier, now Adding column AgeCategory will add if <18 child female Adult if age >18 and sex is female and male Adult if age>18 and sex is male 


```{r}

titanic_df$AgeCategory[titanic_df$Age <= 18] <-'child'
titanic_df$AgeCategory[titanic_df$Age > 18.00 & titanic_df$Sex =='female']<- 'FemaleAdult'
titanic_df$AgeCategory[titanic_df$Age > 18.00 & titanic_df$Sex =='male']<- 'maleAdult'
titanic_df$AgeCategory<-factor(titanic_df$AgeCategory)

#just to verify the data 
datatable(titanic_df, options = list(pageLength = 5))



```

Now all the required data has been filled and resturtured so splitting again the data into training and test data 

```{r}
final_train_df<-titanic_df[1:nrow(train_df),]
final_test_df<-titanic_df[nrow(train_df)+1:nrow(test_df),]

datatable(final_train_df, options = list(pageLength = 5))

datatable(final_test_df, options = list(pageLength = 5))



```

#Predict 
predicting the model on training data 
here i am using decision tree to model 
as we have seen only few variable are significant so considering only with Pclass,Age,Sex,AgeCategory,FamgpSize

```{r}
library("rpart")
library("rpart.plot")
rtree_fit <- rpart(Survived ~ Pclass+Age+factor(AgeCategory)+Sex+groupFmSize, 
          final_train_df,method="class") 
summary(rtree_fit)
rpart.plot(rtree_fit,extra=104, box.palette="GnBu",
branch.lty=3, shadow.col="gray", nn=TRUE)
```
Insights

from the first level of tree shows out of total traveller only 38% survived while 62 % died
and further only 19 % of male survievd however this ratio was more in case of women 

#prediction
```{r}
 library(MASS) 


survivalPrediction <- predict(rtree_fit,final_test_df,type="class")
str(survivalPrediction)

tableP<-table(survivalPrediction)
pct <- round(tableP/sum(tableP) * 100)
label<-c("0","1")
lbls <- paste(label,'-',pct,'%')  # add percents to labels 
 pie(tableP,col=c("red","yellow"),labels = lbls)  
 
 #save the data in CSV file 
predicted_df <- data.frame(PassengerID = final_test_df$PassengerId, Survived = survivalPrediction)
head(predicted_df,200)


write.csv(predicted_df, file = 'Titanic_Prediction_Rpart.csv', row.names = F)




```

#conclusion

From the final data pie chart 65% people has been died while 35% has survived from disaster.


This is my first Exploration .Any improvments and  suggestions always welcome:



